BlockTensorDecompositions.jl: A Unified Constrained Tensor Decomposition Julia Package

### 2025-02-04 paper/paper.qmd
First pass of momentum section

should be thought of as the following entry-wise equation

@@

### Computing Gradients {#sec-gradient-computation}
- Use Auto diff generally
- But hand-crafted gradients and Lipschitz calculations *can* be faster (e.g. symmetrized slicewise dot product)

@@

It is not too hard to show that ${L}_{n,r}^t$ is the Euclidean norm of the $r$th column^[We could have used the $r$th row of $\tilde{X}_n \cdot_n \tilde{X}_n$ since this matrix is symmetric. Since Julia store matrices in column-major order, many operations that perform column-wise are more efficient than their equivalent row-wise operation.] of the matrix $\tilde{X}_n \cdot_n \tilde{X}_n$ from @cor-least-squares-matrix,

@@

$$
\begin{align}
\nabla f_{0}^t(B^t) \times_B (\hat{L}_{B}^t)^{-1} &= \nabla f_{0}^t(B^t)  \bigtimes_{ n} (\hat{L}_{B,n}^{t})^{-1} = \llbracket \nabla f_{0}^t(B^t) ; (\hat{L}_{B,1}^{t})^{-1},\dots,(\hat{L}_{B,n}^{t})^{-1} \rrbracket.
\end{align}
$$

@@

@@

$$
\hat{A}_n^t \leftarrow A_n^t + \omega_n^t (A_n^{t} - A_n^{t-1})
$$

@@

Our selection for $\omega_n^t$ follows Xu and Yin's method for block coordinate descent @xu_BlockCoordinateDescent_2013, which is itself inspired by Tseng and Yun's coordinate gradient descent method @tseng_coordinate_2009.

TODO CONTINUE FROM HERE

```julia
function (U::MomentumUpdate)(x::T; x_last::T, ω, δ, kwargs...) where T
    n = U.n
    if checkfrozen(x, n)
        return x
    end
    # TODO avoid redoing this lipschitz calculation and instead store the previous L
    # TODO generalize this momentum update to allow for other decaying momentums ω
    L = U.lipschitz(x; kwargs...)
    L_last = U.lipschitz(x_last; kwargs...)
    ω = min.(ω, δ .* .√(L_last/L))

    a, a_last = factor(x, n), factor(x_last, n)

    # Equivalent mathematically, but slightly less efficient ways of applying momentum
    # @. a = a + ω * (a - a_last)

    # @. a += ω * (a - a_last)

    # a .*= 1 + a
    # a .-= ω .* a_last

    a = U.combine(a, id + ω) # handle diagonal Lipschitz constants
    a .-= U.combine(a_last, ω)
end
```

```julia
"""update parameters needed for the update"""
function initialize_parameters(decomposition, Y, previous; momentum::Bool, random_order, recursive_random_order, kwargs...)
	# parameters for the update step are symbol => value pairs
	# they are held in a dictionary since we may mutate these for ex. the stepsize
	parameters = Dict{Symbol, Any}()

	# General Looping
	parameters[:iteration] = 0
	parameters[:x_last] = previous[begin] # Last iterate
	parameters[:random_order] = random_order
	parameters[:recursive_random_order] = recursive_random_order

	# Momentum
	if momentum
		parameters[:t_last] = float(1) # need this field to hold Floats, not Ints
		parameters[:t] = update_t(float(1))
		parameters[:ω] = (parameters[:t_last] - 1) / parameters[:t]
		parameters[:δ] = kwargs[:δ]
	end

	function updateparameters!(parameters, decomposition, previous)
		parameters[:iteration] += 1
		#parameters[:x_last] = previous[begin] # Last iterate updated with updateprevious!
											   # this ensures we only copy over the decomposition
											   # once since updateparameters! happens before updateprevious!

											   # This is commented since parameters[:x_last] already points to previous[begin]
		# Momentum
		if momentum
			parameters[:t_last] = parameters[:t]
			parameters[:t] = update_t(parameters[:t_last])
			parameters[:ω] = (parameters[:t_last] - 1) / parameters[:t]
		end
	end

@@

>>>>

in @thm-operator-norm-outer-product should be thought of as the following entry-wise equation

@@

### Computing Gradients {#sec-gradient-computation}
- Use Auto diff generally
- But hand-crafted gradients and Lipschitz calculations *can* be faster (e.g. symmetrized slice-wise dot product)

@@

It is not too hard to show that the Lipschitz ${L}_{n,r}^t$ for $\nabla f_{n,r}^t$ is the Euclidean norm of the $r$th column^[We could have used the $r$th row of $\tilde{X}_n \cdot_n \tilde{X}_n$ since this matrix is symmetric. Since Julia store matrices in column-major order, many operations that perform column-wise are more efficient than their equivalent row-wise operation.] of the matrix $\tilde{X}_n \cdot_n \tilde{X}_n$ from @cor-least-squares-matrix,

@@

$$
\begin{align}
\nabla f_{0}^t(B^t) \times_B (\hat{L}_{B}^t)^{-1} &= \nabla f_{0}^t(B^t)  \bigtimes_{ n} (\hat{L}_{B,n}^{t})^{-1} \\
&= \llbracket \nabla f_{0}^t(B^t) ; (\hat{L}_{B,1}^{t})^{-1},\dots,(\hat{L}_{B,n}^{t})^{-1} \rrbracket.
\end{align}
$$ {#eq-tensor-matrices-product}

@@

TODO use the multi-mode product in stead of defining a new multiplication $\times_B$ here. I think I'll have the same tensor product issue as before where the indices.

@@

$$
\hat{A}_n^t \leftarrow A_n^t + \omega_n^t(A_n^{t} - A_n^{t-1})
$$ {#eq-extrapolation-ideal}

@@

Our selection for $\omega_n^t$ follows Xu and Yin's method for block coordinate descent [@xu_BlockCoordinateDescent_2013], which is itself inspired by Tseng and Yun's coordinate gradient descent method [@tseng_coordinate_2009].

Given a parameter^[Usually we pick a number close to $1$. For example, we use the default $\delta=0.9999$.] $\delta \in [0, 1)$, we define the momentum parameters and $τ^t$ and $\omega_n^t$ according to the following updates

$$
\begin{align*}
\tau^0 &= 1 \\
\tau^{t+1} &\leftarrow \frac{1}{2} \left(1+\sqrt{1+4\left(\tau^{t}\right)^2}\right) \\
\hat{\omega}^{t} &\leftarrow \frac{\tau^{t} - 1}{\tau^{t+1}} \\
\omega_n^{t} &\leftarrow \min\left(\hat{\omega}^{t}, \delta \sqrt{\hat{L}_n^{t-1} (\hat{L}_n^{t})^{-1}} \right).
\end{align*}
$$ {#eq-momentum-parameters}

TODO notation is going to get confusing. We use hat/not hat $L$ for the scaler vs matrix/tensor version. But we use hat/not hat $\omega$ for the ideal vs clamped momentum.

What is novel with our approach is that we perform this momentum on the Lipschitz matrices and tensors $\hat{L}_n^{t}$ rather than scaler Lipschitz constant ${L}_n^{t}$. In this way, we should interpret the operations shown in @eq-momentum-parameters as operating element-wise. This also means the momentum parameter $\omega_n^{t}$ is a matrix or tensor and takes the same shape as $\hat{L}_n^{t}$.

In order to perform @eq-extrapolation-ideal, we use the equivalent but more efficient formulation

$$
\begin{align}
\hat{A}_n^{t} &\leftarrow A_n^t (\text{id}_{R_n} + \omega_n^t) - {A}_n^{t-1} \omega_n^t.
\end{align}
$$

```julia
function (U::MomentumUpdate)(X::T; X_last::T, ω, δ, kwargs...) where T
    n = U.n

    L = U.lipschitz(X; kwargs...)
    L_last = U.lipschitz(X_last; kwargs...)
    ω = min.(ω, δ .* .√(L_last/L))

    A, A_last = factor(X, n), factor(X_last, n)

    A .= U.combine(A, id + ω)
    A .-= U.combine(A_last, ω)
end
```

In the code above, the momentum stores the factor $n$ in acts on, how to compute the Lipschitz constant, matrix, or tensor, and how to combine (multiply) the constant with the factor. In the case of matrix factors $A_n$ in a Tucker decomposition, this is simply right matrix-matrix multiplication. The core factor $B$ uses $\times_B$ as described in @eq-tensor-matrices-product.

The parameters of the momentum update are handled separately. This is to treat the momentum update as "apply this update with these parameters". The parameters $\tau^t$ and $\hat{\omega}^t$ are updated by the following function that keeps track of all parameters needed to perform the iteration. In this case, we keep track of what iteration $t$ we are at, the previous iterate, and a few options for the order in which to cycle through and update the blocks.

```julia
update_τ(τ) = 0.5*(1 + sqrt(1 + 4*τ^2))

function initialize_parameters(decomposition, Y, previous; momentum::Bool, random_order, recursive_random_order, kwargs...)
	# parameters for the update step are symbol => value pairs
	# held in a dictionary since we may mutate these, e.g. the step-size
	parameters = Dict{Symbol, Any}()

	# General Looping
	parameters[:iteration] = 0
	parameters[:X_last] = previous[begin] # Last iterate
	parameters[:random_order] = random_order
	parameters[:recursive_random_order] = recursive_random_order

	# Momentum
	if momentum
		parameters[:τ_last] = float(1) # need this field to hold Floats, not Ints
		parameters[:τ] = update_τ(float(1))
		parameters[:ω] = (parameters[:τ_last] - 1) / parameters[:τ]
		parameters[:δ] = kwargs[:δ]
	end

	function updateparameters!(parameters, decomposition, previous)
		parameters[:iteration] += 1
		# parameters[:x_last] = previous[begin]
# This is commented since parameters[:x_last] already points to previous[begin]

		if momentum
			parameters[:τ_last] = parameters[:τ]
			parameters[:τ] = update_τ(parameters[:τ_last])
			parameters[:ω] = (parameters[:τ_last] - 1) / parameters[:τ]
		end
	end

@@

### Empirical evidence for
